---
title: "Reproducible Research: Peer Assessment 1"
output:   html_document
keep_md: true
---


## Loading and preprocessing the data
```{r load and preprocess}
    library("reshape2")
    df<-read.csv("activity.csv",header=TRUE, stringsAsFactors=FALSE)
    raw<-df
    data<-na.omit(df)
    data$steps <- as.numeric(data$steps)
    data$interval <- as.numeric(data$interval)
    melted <- melt(data,id=c(2:3),measure=c(1),na.rm=TRUE)
    sumsteps <- dcast(melted, date~variable,sum)
```  


## What is mean total number of steps taken per day?
```{r mean steps}
    meansteps <- mean(sumsteps$steps)
    print(meansteps)
```


## What is the average daily activity pattern?
```{r median steps}
    mediansteps <- median(sumsteps$steps)
    print(mediansteps)
```  

####Histogram
```{r histogram,fig.height=4, fig.width=4,echo=FALSE}
    #One for the output
    hist(sumsteps$steps,xlab="Total Number of Daily Steps",
         main="Total number of steps taken each day")
    #Generate a plot that can be checked into github or attached to some
    #other document...
    png(file="hist.png",height=480,width=480)
    hist(sumsteps$steps,xlab="Total Number of Daily Steps",
         main="Total number of steps taken each day")
    dev.off()
```  

##Average Daily Activity Pattern
```{r timeseries}
    #Calculate the average for each 5 minute interval over all the days
    meantimeseries <- dcast(melted,interval~variable,mean)

    #Create a separate figure that can be checked into github or attached to
    #another document...
    png(file="timeseries.png", height=480, width=480)
    plot(meantimeseries$interval,
         meantimeseries$steps,type="l",
         xlab="Time Interval (seconds)",
         ylab="Avg # steps across all days")
    dev.off()

    #Create a figure that is viewed in the knitr output
    plot(meantimeseries$interval,
         meantimeseries$steps,type="l",
         xlab="Time Interval (seconds)",
         ylab="Avg # steps across all days")
    
    
```  

###Time interval with the maximum number of steps (averaged across all days):
```{r time interval with max steps}
    index<-which.max(meantimeseries$steps)
    print(cat( meantimeseries[index,1], " second interval"))
```  

## Imputing missing values

####1.  Total number of missing values (NA) in dataset
```{r number of NAs}
    num.missing <- as.numeric(sum(is.na(df$steps)))
    print("Number of NA's for step data: ")
    print(num.missing)
```  

####2.  Strategy for filling in the missing values
Replace each *NA* with the mean value that corresponds to the 5-minute interval
of the missing value.  The mean value for each 5-minute interval is available
from the time-series calculation we performed earlier.  


####3.  Create a new dataset with missing values filled in with the median value
```{r fill missing data}
    missingindices <- which(is.na(raw$steps))
    rawdata <- df
    #Replace each NA with the average number of steps for the interval that
    #corresponds to this.
    for(i in 1:length(missingindices)){
       #get the time interval for each element in the missingindices list/vector
       rawinterval <- rawdata[i,3]

       #get the mean number of steps for this interval (from the
       #meantimeseries data frame)
       row <- which(meantimeseries$interval == rawinterval)
       meanforinterval<- meantimeseries[row,2]

       #Replace the NA for this row with the meanforinterval value
       rawdata[missingindices[i],1] <- meanforinterval

   }
  #print(head(rawdata,500))

```  

####4. Make histogram, calculate and report the mean and median total number of steps taken per day.

####Histogram- Filled Data
```{r histogram filled data,fig.height=4, fig.width=4,echo=FALSE}
    melted <- melt(rawdata,id=c(2:3),measure=c(1),na.rm=TRUE)
    sumsteps2 <- dcast(melted, date~variable,sum)
    #One for the output
    hist(sumsteps2$steps,xlab="Total Number of Daily Steps",
         main="Total number of steps taken each day")
    #Generate a plot that can be checked into github or attached to some
    #other document...
    png(file="filledhist.png",height=480,width=480)
    hist(sumsteps2$steps,xlab="Total Number of Daily Steps",
         main="Total number of steps taken each day (filled data")
    dev.off()
```  
####Comments:
Replacing the missing data with the average number of steps (that correspond to
the time interval) increased the frequency of the number of averages steps but
did not change the "shape" of the histogram.


####Filled Data-Mean total number of steps taken per day:
```{r mean steps-filled data}
    meansteps2 <- mean(sumsteps2$steps)
    print(meansteps2)
```

####Filled Data-Median number of total daily steps:
```{r median steps- filled data}
    mediansteps2 <- median(sumsteps2$steps)
    print(mediansteps2)
```


## Are there differences in activity patterns between weekdays and weekends?
```{r days of week,echo=FALSE, fig.width=4, fig.height=8 }
    rawdata$date <- as.Date(rawdata$date)
    wday <- weekdays(rawdata$date)
    for(i in 1:length(wday)){
       if (wday[i] =="Sunday" || wday[i] =="Saturday"){
           wday[i] = "weekend"
       }
       else{
          wday[i] = "weekday"
      }
    }
    dayofweek <- as.factor(wday)

    rawdata$dayofweek <-cbind(c(wday))
    weekdays.data <- subset(rawdata,dayofweek=="weekday")
    weekends.data <- subset(rawdata,dayofweek=="weekend")
    melted.weekday <- melt(weekdays.data,id=c(2:4),measure=c(1),na.rm=FALSE)
    melted.weekend <- melt(weekends.data, id=c(2:4),measure=c(1),na.rm=FALSE)
    meantimeseries.weekday <- dcast(melted.weekday,interval~variable,mean)
    meantimeseries.weekend <- dcast(melted.weekend,interval~variable,mean)
    png(file="dayofweek.png", height=480, width=480)
    par(mfrow=c(2,1))
    plot(meantimeseries.weekday$interval,
         meantimeseries.weekday$step,type="l",
         xlab="Time Interval (seconds)",
         ylab="Number of steps",
         main="Weekdays")

    plot(meantimeseries.weekend$interval,
         meantimeseries.weekend$step,type="l",
         xlab="Time Interval (seconds)",
         ylab="Number of steps",
         main="Weekends")

   dev.off()

   #Generate the panel plot that is viewable from knitr output
   par(mfrow=c(2,1))
   plot(meantimeseries.weekday$interval,
        meantimeseries.weekday$step,type="l",
        xlab="Time Interval (seconds)",
        ylab="Number of steps",
        main="Weekdays")

  plot(meantimeseries.weekend$interval,
        meantimeseries.weekend$step,type="l",
        xlab="Time Interval (seconds)",
        ylab="Number of steps",
        main="Weekends")

```  

####Comments:
There are more steps taken throughout the day on weekends, especially from the
1000 sec to ~1800 sec interval.  The number of steps taken are also taken later
on the weekend mornings, consistent with sleeping-in on the weekends.



